---
title: "Computational Text Analysis - Exercise 3: Comparison and complexity"
author: "Leung Ka Hei"
date: "25/02/2026"
output: html_document
---

## Introduction

The hands-on exercise for this week focuses on: 1) comparing texts; 2) measuring he document-level characteristics of text---here, complexity.

In this exercise, you will learn how to:
  
* Compare texts using character-based measures of similarity and distance
* Compare texts using term-based measures of similarity and distance
* Calculate the complexity of texts
* Replicate analyses from @schoonvelde_liberals_2019

## Setup 

Before proceeding, we'll load the remaining packages we will need for this exercise.

```{r, echo=F}
library(kableExtra)
```

```{r, message=F}
library(readr) # more informative and easy way to import data
library(quanteda) # includes functions to implement Lexicoder
library(quanteda.textstats) # for estimating similarity and complexity measures
library(stringdist) # for basic character-based distance measures
library(dplyr) #for wrangling data
library(tibble) #for wrangling data
library(ggplot2) #for visualization
```


For this example we'll be using data from the 2017-2018 Theresa May Cabinet in the UK. The data are tweets by members of this cabinet. 

```{r}
tweets  <- readRDS(gzcon(url("https://github.com/cjbarrie/CTA-ED/blob/main/data/comparison-complexity/cabinet_tweets.rds?raw=true")))
```

And we see that the data contain three variables: "username," which is the username of the MP in question; "tweet," which is the text of the given tweet, and "date" in days in yyyy-mm-dd format. 

```{r}
head(tweets)
```

And there are 24 MPs whose tweets we're examining. 

```{r}
unique(tweets$username)

length(unique(tweets$username))
```

## Generate document feature matrix

In order to use the `quanteda` package and its accompanying `quanteda.textstats` package, we need to reformat the data into a quanteda "corpus" object. To do this we just need to specify the text we're interested in as well as any associated document-level variables in which we're interested. 

We can do this as follows. 

```{r}
#make corpus object, specifying tweet as text field
tweets_corpus <- corpus(tweets, text_field = "tweet")

#add in username document-level information
docvars(tweets_corpus, "username") <- tweets$username

tweets_corpus
```

We are now ready to reformat the data into a document feature matrix.

```{r}
dfmat <- dfm(tokens(tweets_corpus, remove_punct = TRUE)) %>%
  dfm_remove(pattern = stopwords("english")) #tells quanteda to remove all English stopwords

dfmat
```

Note that when we do this we need to have tokenized our corpus object first. We can do this by wrapping the `tokens` function inside the `dfm()` function as above. 

So what is this object? Well the documents here are tweets. And the matrix is a sparse (i.e., mostly zeroes) matrix of 1s and 0s for whether a given word appears in the document (tweet) in question. 

The vertical elements (columns) of this vector are made up of all the words used in all of the tweets combined. Here, it helps to imagine every tweet positioned side by side to understand what's going on here. 

## Compare between MPs

Once we have our data in this format, we are ready to compare between the text produced by members of Theresa May's Cabinet.

Here's an example of the correlations between the combined tweets of 5 of the MPs with each other.

```{r}
corrmat <- dfmat %>%
  dfm_group(groups = username) %>% #groupby+summarise
  textstat_simil(margin = "documents", method = "correlation") #computes how similar each document is to every other document, using correlation as the similarity measure

#correlation: same word? similar proportion? same vocabulary space?

corrmat[1:5,1:5]

```

Note that here we're using the `dfm_group()` function, which allows you to take a document feature matrix and make calculations while grouping by one of the document-level variables we specified above. 

There are many different measures of similarity, however, that we might think about using.

In the below, we combine four different measures of similarity, and see how they compare to each other across MPs. Note that here we're looking only at the similarity between an MP's tweets and those of then Prime Minister, Theresa May.

## Compare between measures

Let's see what this looks like for one of these measures---cosine similarity.

We first get similarities between the text of MP tweets and all other MPs.

```{r}

#estimate similarity, grouping by username
#cosine-> vocabulary overlap and direction, not about: style, tone, abstract or not, which voice, distributional shape, patterns

cos_sim <- dfmat %>%
  dfm_group(groups = username) %>%
  textstat_simil(margin = "documents", method = "cosine") #specify method here as character object

```

But remember we're only interested in how they compare to what Theresa May has been saying. 

So we need to take these cosine similarities and retain only those similarity measures corresponding to the text of Theresa May's tweets. 

We first convert the `textstat_simil()` output to a matrix.

```{r}

cosmat <- as.matrix(cos_sim) #convert to a matrix

head(cosmat)
```

And we can see that the 23rd row of this matrix contains the similarity measures with the Theresa May tweets. 

We take this row, removing the similarity of Theresa May with herself (which will always = 1), and convert it to a datframe object. 

```{r}
#generate data frame keeping only the row for Theresa May
cosmatdf <- as.data.frame(cosmat[23, c(1:22, 24)]) #keeps row correct in meaning 23 but drops column 23, leaving only columns 1–22 and 24
```

We then rename the cosine similarity column with an appropriate name and convert row names to a column variable so that we have cells containing information on the MP to which the cosine similarity measure refers. 

```{r}
#rename column
colnames(cosmatdf) <- "corr_may"
  
#create column variable from rownames
cosmatdf <- tibble::rownames_to_column(cosmatdf, "username") #tibble package -> takes row names 1, 2,3,4 turn them to column called usename
```

And like so we have our data in tidy format, which we can then plot like so. 

```{r}
ggplot(cosmatdf) +
  geom_point(aes(x=reorder(username, -corr_may), y= corr_may)) + 
  coord_flip() +
  xlab("MP username") +
  ylab("Cosine similarity score") + 
  theme_minimal()

```

Combining these steps into a single `for` loop, we can see how our different similarity measures of interest compare. 

```{r}

#specify different similarity measures to explore
methods <- c("correlation", "cosine", "dice", "edice")

#correlation: compares patterns of word usage -> style, distribution, or relative emphasis, patterns esp.
#consine: compares relative proportions of words, so it detects ideology only when ideology consistent vocabulary is expressed through choices
#dice: only at whether a word is present or absent, and then measures how much the two documents’ sets of unique words overlap
#edice: counting how much the two documents’ word‑frequency amounts overlap

#create empty dataframe
testdf_all <- data.frame()

#gen for loop across methods types
for (i in seq_along(methods)) { 
  
  #pass method to character string object
  sim_method <- methods[[i]] #i = 1 runs the code using "correlation", i = 2 runs it using "cosine", and the loop continues the same way for "dice" and "edice"
  
  #estimate similarity, grouping by username
  test <- dfmat %>%
    dfm_group(groups = username) %>%
    textstat_simil(margin = "documents", method = sim_method) #specify method here as character object created above
  
  testm <- as.matrix(test) #convert to a matrix
  
  #generate data frame keeping only the row for Theresa May
  testdf <- as.data.frame(testm[23, c(1:22, 24)])
  
  #rename column
  colnames(testdf) <- "corr_may"
  
  #create column variable from rownames
  testdf <- tibble::rownames_to_column(testdf, "username")
  
  #record method in new column variable
  testdf$method <- sim_method

  #bind all together
  testdf_all <- rbind(testdf_all, testdf)  
  
}

#create variable (for viz only) that is mean of similarity scores for each MP
testdf_all <- testdf_all %>%
  group_by(username) %>%
  mutate(mean_sim = mean(corr_may))

ggplot(testdf_all) +
  geom_point( aes(x=reorder(username, -mean_sim), y= corr_may, color = method)) + 
  coord_flip() +
  xlab("MP username") +
  ylab("Similarity score") + 
  theme_minimal()

```

## Complexity

We now move to document-level measures of text characteristics. And here we will focus on the paper by @schoonvelde_liberals_2019. 

We will be using a subset of these data, taken from EU speeches given by four politicians. These are provided by the authors at [https://dataverse.harvard.edu/dataset.xhtml?persistentId=doi:10.7910/DVN/S4IZ8K](https://dataverse.harvard.edu/dataset.xhtml?persistentId=doi:10.7910/DVN/S4IZ8K).

We can load the data as follows.

```{r}

speeches  <- readRDS(gzcon(url("https://github.com/cjbarrie/CTA-ED/blob/main/data/comparison-complexity/speeches.rds?raw=true")))

```

And we can take a look at what the data contains below.

```{r}

head(speeches)

```

The data contain speeches by four different politicians, each of whom are positioned at different points on a liberal-conservative scale.

We can then calculate the Flesch-Kincaid readability/complexity score with the `quanteda.textstats` package like so.

```{r}
speeches$flesch.kincaid <- textstat_readability(speeches$text, measure = "Flesch.Kincaid") #Calculate the readability score of each document using the Flesch–Kincaid formula： produces a grade‑level number indicating the years of education typically needed to comprehend the text

# returned as quanteda data.frame with document-level information;
# need just the score:
speeches$flesch.kincaid <- speeches$flesch.kincaid$Flesch.Kincaid
```

We want this information aggregated over each of our politicians: Gordon Brown, Jose Zapatero", David Cameron, and Mariano Rajoy. These are recorded in the data under a column called "speaker."

```{r}
#get mean and standard deviation of Flesch-Kincaid, and N of speeches for each speaker
sum_corpus <- speeches %>%
  group_by(speaker) %>%
  summarise(mean = mean(flesch.kincaid, na.rm=TRUE),
                   SD=sd(flesch.kincaid, na.rm=TRUE),
                   N=length(speaker))

# calculate standard errors and confidence intervals
#to see how trustworthy each speaker’s average readability score is
sum_corpus$se <- sum_corpus$SD / sqrt(sum_corpus$N)
sum_corpus$min <- sum_corpus$mean - 1.96*sum_corpus$se
sum_corpus$max <- sum_corpus$mean + 1.96*sum_corpus$se
```

And this gives us data in tidy format that looks like so. 

```{r}
sum_corpus
```

Which we can then plot---and we see that our results look like those in Figure 1 of the published article by @schoonvelde_liberals_2019. 


```{r}

ggplot(sum_corpus, aes(x=speaker, y=mean)) +
  geom_bar(stat="identity") + 
  geom_errorbar(ymin=sum_corpus$min,ymax=sum_corpus$max, width=.2) +
  coord_flip() +
  xlab("") +
  ylab("Mean Complexity") + 
  theme_minimal() + 
  ylim(c(0,20))

```

## Exercises

1. Compute distance measures such as "euclidean" or "manhattan" for the MP tweets as above, comparing between tweets by MPs and tweets by PM, Theresa May. 
```{r}
dfmat_ <- dfm(tokens(tweets_corpus, remove_punct = TRUE)) %>%
  dfm_remove(pattern = stopwords("english"))

#group tweets by username
dfmat_grouped <- dfm_group(dfmat_, groups = username)

#calculating distance: Euclidean
dist_euc <- textstat_dist(dfmat_grouped, method = "euclidean")

#calculating distance: Manhattan
dist_man <- textstat_dist(dfmat_grouped, method = "manhattan")

#find Theresa May's row
rownames(dfmat_grouped)

#convert to matrix
mat_euc <- as.matrix(dist_euc) 
mat_man <- as.matrix(dist_man)

#generate data frame keeping only the row for Theresa May
euc_df <- as.data.frame(mat_euc[23, c(1:22, 24)])
man_df <- as.data.frame(mat_man[23, c(1:22, 24)])

#rename column
colnames(euc_df) <- "corr_may"
colnames(man_df) <- "corr_may"

#create column variable from rownames
euc_df <- tibble::rownames_to_column(euc_df, "username")
man_df <- tibble::rownames_to_column(man_df, "username")

#plot
ggplot(euc_df) +
  geom_point(aes(x=reorder(username, -corr_may), y= corr_may)) + 
  coord_flip() +
  xlab("MP username") +
  ylab("Euclidean distance score") + 
  theme_minimal()

ggplot(man_df) +
  geom_point(aes(x=reorder(username, -corr_may), y= corr_may)) + 
  coord_flip() +
  xlab("MP username") +
  ylab("Manhattan distance score") + 
  theme_minimal()

#combining into a single step by loop
#specify distance measure
methods_ <- c("euclidean", "manhattan")

#create empty dataframe for bind the results of two methods together
testdf_both <- data.frame()

#gen for loop across methods types
for (i in seq_along(methods_)) {
  
  #pass method to character string object
  dis_method <- methods_[[i]]
  
  #estimate distance, grouping by username
  test_ <- dfmat %>%
    dfm_group(groups = username) %>%
    textstat_dist(margin = "documents", method = dis_method) #specify method here as character object created above
  
  test_mat <- as.matrix(test_) #convert to a matrix
  
  #generate data frame keeping only the row for Theresa May
  testdf <- as.data.frame(test_mat[23, c(1:22, 24)])
  
  #rename column
  colnames(testdf) <- "corr_may"
  
  #create column variable from rownames
  testdf <- tibble::rownames_to_column(testdf, "username")
  
  #record method in new column variable
  testdf$method <- dis_method

  #bind all together
  testdf_both <- rbind(testdf_both, testdf)  
  
}

#create variable that is the mean of distance scores for each MP
testdf_both <- testdf_both %>%
  group_by(username) %>%
  mutate(mean_dis = mean(corr_may))

#plot
ggplot(testdf_both) +
  geom_point( aes(x=reorder(username, -mean_dis), y= corr_may, color = method)) + 
  coord_flip() +
  xlab("MP username") +
  ylab("Distance score") + 
  theme_minimal()
```
The plot shows how closely each individual MP’s tweeting style aligns with Theresa May’s by comparing their document‑feature vectors using Euclidean and Manhattan distance. The red dots show the Euclidean distance between each MP’s tweets and Theresa May’s tweets, while the blue dots show the corresponding Manhattan distance. Since the distance reflects dissimilarity, MPs with the greater scores meaning use language that diverges more from Theresa May's tweets, and those who with smaller scores meaning use language more similar to May's tweets in vocabulary and thematic emphasis. In the plot, Manhattan distances are larger compared to Euclidean distances, yet the overall ordering of MPs remain stable. This indicates that the choice of metric does not change the substantive interpretation.

From the plot, MPs such as daviddavismp and philiphammonduk have distance score nearly to zero, ndicating that their tweeting style is highly similar to Theresa May’s. Meanwhile, MPs such as brandonlewis and pennymordanut score over 15,000 and 20,000 in Manhattan distance scale, reflecting that they have more distinct communication priorities or stylistic choices from Theresa May's. Overall,  the plot highlights meaningful variation within the Conservative Party, with some MPs echoing May’s themes more closely than others.

2. Estimate at least three other complexity measures for the EU speeches as above. Consider how the results compare to the Flesch-Kincaid measure used in the article by @schoonvelde_liberals_2019.
```{r}
#load data
speeches_ <- readRDS(gzcon(url("https://github.com/cjbarrie/CTA-ED/blob/main/data/comparison-complexity/speeches.rds?raw=true")))

head(speeches_)

#compute the baseline
fk <- textstat_readability(speeches_$text, measure = "Flesch.Kincaid")
speeches_$flesch.kincaid <- fk$Flesch.Kincaid

#mean and SE
sum_corpus <- speeches_ %>%
  group_by(speaker) %>%
  summarise(
    mean = mean(flesch.kincaid, na.rm = TRUE),
    SD = sd(flesch.kincaid, na.rm = TRUE),
    N = length(speaker)
  )

sum_corpus$se <- sum_corpus$SD / sqrt(sum_corpus$N)
sum_corpus$min <- sum_corpus$mean - 1.96 * sum_corpus$se
sum_corpus$max <- sum_corpus$mean + 1.96 * sum_corpus$se

#computing other three methods
others <- textstat_readability(
  speeches_$text,
  measure = c("Flesch", "FOG", "Coleman.Liau", "SMOG")
)

speeches_$flesch <- others$Flesch
speeches_$fog <- others$FOG
speeches_$coleman <- others$Coleman.Liau
speeches_$smog <- others$SMOG

#summaries with one loop
measures <- c("flesch.kincaid", "flesch", "fog", "coleman", "smog")

summary_list <- list()

for (m in measures) {
  
  tmp <- speeches_ %>%
    group_by(speaker) %>%
    summarise(
      mean = mean(.data[[m]], na.rm = TRUE),
      SD   = sd(.data[[m]], na.rm = TRUE),
      N    = n()
    ) %>%
    mutate(
      se  = SD / sqrt(N),
      min = mean - 1.96 * se,
      max = mean + 1.96 * se,
      measure = m
    )
  
  summary_list[[m]] <- tmp
}

sum_all <- bind_rows(summary_list)

#plot
ggplot(sum_all, aes(x = reorder(speaker, -mean), 
                    y = mean, 
                    colour = measure)) +
  geom_pointrange(aes(ymin = min, ymax = max),
                  position = position_dodge(width = 0.6)) +
  coord_flip() +
  theme_minimal() +
  xlab("Speaker") +
  ylab("Mean readability score (95% CI)")


```
The figure compares the readability of speeches delivered by Zapatero, Rajoy, Brown, and Cameron across five widely used readability formulas. While the five readability formula in the plot operate on different scales, it means a 'higher' or 'lower' score does not always correspond to the same interpretation across measures. For Flesch Ease and Coleman-Liau, higher values indicating easier-to-read text, conversely, for Flesch-kincaid, Gunning Fog, and SMOG, lower values indicating easier text for reading. Since each formula captures a different dimension of linguistic complexity, including sentence length, word length, and proportion of complex words, the relative positions of the four speakers must be interpreted within each metric rather than compared directly across them.

Even though the five readability formulas operate on different scales, the overall pattern across speakers is broadly consistent. Zapatero receives the lowerst in both Flesch Ease and Coleman score with approximately 40 respectively, and the highest scores on Flesch-kincaid, Gunning Fog, and SMOG between 15-20 each, revealing that his speeches are the hardest to read in both sentence structure and vocabulary. Cameron, on the contrary, score highest on the Flesch Ease and Coleman-Liau scale on approximately 50-55 perspectively, and lowest scores in Flesch-kincaid, Gunning Fog, and SMOG that all below 15. This pattern indicates that Cameron's speeches are the easiest to read overall, combining relatively simple sentence structure with less complex vocabulary. Rajoy and Brown generally falls between these extremes. 

Together, these results highlight stable stylistic differences among the four leaders, with Zapatero combining dense sentence structure with simple vocabulary, and Cameron combining simpler syntax with more complex word choice.

3. (Advanced---optional) Estimate similarity scores between the MP tweets and the PM tweets for each week contained in the data. Plot the results. 
```{r}
#Create the PM/MP role variable
tweets <- tweets %>%
  mutate(role = ifelse(username == "theresa_may", "PM", "MP"))

#turn date to week 
tweets <- tweets %>%
  mutate(date = as.Date(date)) %>%
  arrange(date) %>%
  mutate(week = as.numeric(format(date, "%U"))) #Convert the date into a Sunday‑based week number, where early January days are week 0

#combine tweets per week and username
weekly_text <- tweets %>%
  group_by(week, role) %>%   # role = PM or MP, grouping all the MPs tgt
  summarise(text = paste(tweet, collapse = " "), .groups = "drop") #collapsing many tweets into one document and removing the grouping structure, so all the tweets from MPs are grouped

#tokenise
weekly_dfm <- weekly_text %>%
  corpus(text_field = "text") %>%
  tokens(remove_punct = TRUE, remove_numbers = TRUE) %>%
  dfm()

#compute similarity per week
weeks <- sort(unique(weekly_text$week))
results <- data.frame()

for (w in weeks) {
  
  # subset dfm to that week
  dfm_w <- weekly_dfm[docvars(weekly_dfm, "week") == w, ]
  
  # compute similarity between all usernames in that week
  sim_w <- textstat_simil(dfm_w, margin = "documents", method = "cosine")
  
  # differetiate PM and MPs
  pm_row <- which(docvars(dfm_w, "role") == "PM") 
  mp_row <- which(docvars(dfm_w, "role") == "MP")
  
  # average similarity to MPs for that week
  if (length(pm_row) == 1 && length(mp_row) == 1) { 
    results <- rbind(results, data.frame( 
    week = w, 
    similarity = sim_w[pm_row, mp_row] #Take the cosine similarity between the PM’s weekly document and the MPs’ weekly document
    ))
  }
}

#plot 
ggplot(results, aes(x = week, y = similarity)) +
  geom_line(size = 1.2, colour = "steelblue") +
  geom_point(size = 2) +
  scale_x_continuous(breaks = c(0, 10, 20, 30, 40, 50)) +
  theme_minimal() +
  xlab("Week") +
  ylab("Similarity between PM and MPs (cosine)") +
  ggtitle("Weekly similarity between MP and PM tweets")

```
The plot displays the weekly cosine similarity between tweets posted by MPs and those by the Prime Minister, illustrating how closely their communication or content aligns over time based on topics and keywords. Cosine similarity ranges from 0 to 1, with higher values indicating that MPs and the PM are discussing more similar topics. 

Across the most of observed period (from 2017-06-08 to 2018-01-08), the similarity remains high, which ranges between 0.6 to 0.8. It indicates that there was a stable level of alignment in political messaging. A clear deviation occurs around week 20, where similarity drops to roughly 0.5 and even below in around week 30, which indicates a temporary divergence in communication priorities. It reflects there may be political events or issues that prompted MPs to different themes from the PM. Following this dip, similarity rises sharply and reaches the highest point in week 50 at approximately 0.9, which suggests a period that MPs and the PM were emphasising very similar topics during this period. However, their topics became dissimilar right after the peak, and dropped to below 0.6 again.

Overall, the plot highlights both the general consistency of MP–PM topic alignment and the specific weeks where their communication diverges.
